---
title: "Plots"
description: "For _The plausibility of sampling as an algorithmic theory of sentence processing_, Open Mind, 2023"
author: "Jacob Louis Hoover"
date: 2023-05-15
toc: true
format: 
  html:
    code-fold: true
    code-tools: true
  pdf: 
    fontsize: 10pt
    monofont: Latin Modern Mono
    monofontoptions: 
      - Scale=0.75
---

# GAM plots

::: {.callout-important icon="false" collapse="false"}
### Dependencies

- external dependencies (R files, will be sourced)

      ./utils/prepare_data.R
      ./utils/my_pacf_plot.R

- data files

      ./data/surprisal_densities.csv
      ./data/surprisal_means.csv
      ./data/GAM_predictions/data_to_plot_NEW_var0_gauss_wprev_N200.csv
      ./data/GAM_predictions/data_to_plot_NEW_var0_gauss_w3prev_N200.csv
      ./data/GAM_predictions/data_to_plot_NEW_lin0var_wprev_N200.csv
      ./data/GAM_predictions/data_to_plot_NEW_linear_wprevm_N200.csv
      ./data/GAM_predictions/data_to_plot_NEW_wprevm_N200.csv
      ./data/GAM_predictions/data_to_plot_NEW_datasubsetless6_wprevm_N200.csv
      ./data/GAM_predictions/data_to_plot_NEW_datasubsetless12_wprevm_N200.csv
      ./data/model-summaries-NEW-linear-wprevm.rds

- model training data

      ./data/surps_lms.RTs.perword.rds
      ./data/surps_lms.RTs.rds

- model file

      ./data/GAM_models/GPT3-davinci.rds

:::

```{r global_options}
#| code-fold: show
#| code-summary: options

# Where data lives (plot data from GAM fits, etc)
datadir <- "./data/"

# Where to put generated plots
figsdir <- "./figures/"
```

## Setup 

```{r setup}
#| message: false
#| code-fold: true
#| code-summary: Setup

# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(dplyr)
library(magrittr)
library(stringr)
library(ggplot2)
library(patchwork)

theme_set(theme_minimal())
theme_update(plot.title.position = "plot")

## Names
context_len_labs <- c(
  max = "maximum",
  `80` = "80 words",
  bysent = "within sent."
)
est_type_labs <- c(
  mean = "mean",
  var = "variance"
)
term_labs <- c(
  surprisal = "surprisal (on mean)",
  `surprisal.1` = "surprisal (on log sd)"
)
model_labs <- c(
  boyce_ngram = "n-gram",
  boyce_grnn = "LSTM",
  "TXL" = "TXL",
  "GPT2" = "GPT2",
  "GPT2-large" = "GPT2-large",
  "GPT2-xl" = "GPT2-xl",
  "GPT-Neo" = "GPT-Neo",
  "GPT-J" = "GPT-J",
  "GPT3-ada" = "GPT3-ada",
  "GPT3-curie" = "GPT3-curie",
  "GPT3-davinci" = "GPT3-davinci"
)
lm_names <- list(
  "boyce_ngram", "boyce_grnn",
  "TXL",
  "GPT2",
  "GPT2-large",
  "GPT2-xl",
  "GPT-Neo",
  "GPT-J",
  "GPT3-ada",
  "GPT3-curie",
  "GPT3-davinci"
) %>% set_names(., .)

## Utils

separate_modelnames <- function(df, modelnames_colname = model, names = lm_names) {
  df %>%
    tidyr::separate({{modelnames_colname}},
             into = c("model", "by"), sep = "_", extra = "merge") %>%
    mutate(model = ifelse(model == "boyce", paste0("boyce_", by), model)) %>%
    mutate(context_len = ifelse(
      is.na(by), "max", ifelse(stringr::str_starts(model, "boyce"), "bysent", by))
    ) %>% select(-by) %>%
    mutate(context_len = forcats::fct_relevel(context_len, c("max")),
           model = forcats::fct_relevel(model, lm_names))
}
# for nicer plot labels
my_labeller <- labeller(
  context_len = context_len_labs,
  term = term_labs,
  model = model_labs,
  est_type=c(mean="mean RT", var=glue::glue("log(sd RT)")))
```

```{r}
#| code-fold: show
#| code-summary: Options#| 

# Whether to only plot two context lengths or all three
twocontexts <- TRUE # (re-run notebook after modifying)

if (twocontexts) {
  context_lens <- c("max", "bysent")
  context_pref <- "2"
} else {
  context_lens <- c("max", "80", "bysent")
  context_pref <- ""
}
```



## GAM smooths

Plotting the fits of GAM smooths for effect of surprisal on mean and variance in RT.

### Load data

```{r load_data_to_plot}
#| warning: false
#| code-fold: true

load_in_data<- function(f, with_est_types = TRUE) {
  d <- readr::read_csv(f,
    col_names = TRUE, show_col_types = FALSE) %>%
    mutate(context_len = forcats::fct_relevel(context_len, c("max")),
           model = forcats::fct_relevel(model, lm_names))
  if (with_est_types) d <- d %>% mutate(est_type = factor(est_type))
  return(d)
}
# get surprisal means and densities 
# (from natural-stories-surprisals/EDA_surps_SPRT.Rmd)

surprisal_densities <- load_in_data(
  paste0(datadir, "surprisal_densities.csv"), with_est_types = F) %>%
  # there is some odd point at ~37 this makes an unnecessarily long tail
  filter(model %in% lm_names, surprisal < 25) 

surprisal_means <- load_in_data(
  paste0(datadir, "surprisal_means.csv"), with_est_types = F) %>% 
  filter(model %in% lm_names)

models_in_desc_PPL_order <- surprisal_means %>% 
  filter(context_len == "max" | model %>% stringr::str_starts("boyce")) %>% 
  arrange(desc(mean_surprisal)) %>% pull(model) %>% as.vector

# GAMs fit for paper (using fitdata=surps_lms.RTs.rds, and correct lagging)

# smooth0var: nonlinear gaussian models with constant variance
data_to_plot_var0_gauss_wprev <- load_in_data(
  paste0(datadir, "GAM_predictions/data_to_plot_NEW_var0_gauss_wprev_N200.csv"))
data_to_plot_var0_gauss_w3prev <- load_in_data(
  paste0(datadir, "GAM_predictions/data_to_plot_NEW_var0_gauss_w3prev_N200.csv"))

##### Linear control with constant variance, with previous, with family=gaulss()
data_to_plot_lin0var_wprev <- load_in_data(
  paste0(datadir, "GAM_predictions/data_to_plot_NEW_lin0var_wprev_N200.csv"))
##### Linear control with constant variance, with previous, with family=gaulss()
data_to_plot_linear_wprevm <- load_in_data(
  paste0(datadir, "GAM_predictions/data_to_plot_NEW_linear_wprevm_N200.csv"))
##### Main models: Smooth GAMs with previous surprisal predictors for mean only
data_to_plot_smooth_wprevm <- load_in_data(
  paste0(datadir, "GAM_predictions/data_to_plot_NEW_wprevm_N200.csv"))

#### As above (Smooth GAMs with previous surprisal for mean), 
#### but removing most surprising items
# Note, only for GPT-3 Davinci full-context.

data_to_plot_datasubsetless6_smooth_wprevm  <- load_in_data(
  paste0(datadir, "GAM_predictions/data_to_plot_NEW_datasubsetless6_wprevm_N200.csv")) %>%
  mutate(cutoff = 6)
data_to_plot_datasubsetless12_smooth_wprevm <- load_in_data(
  paste0(datadir, "GAM_predictions/data_to_plot_NEW_datasubsetless12_wprevm_N200.csv")) %>%
  mutate(cutoff = 12)
```



### Plot smooths


```{r}
#| code-fold: true
#| warning: false

plot_data <- function(
    data_to_plot, use_shifted_est=TRUE,
    se.mult=1.96, # use approx qnorm(1 - 0.05/2)
    trans = I, #transform function to apply to the yvalues
    controls = NULL,
    surp_units = "nats",
    density_df = surprisal_densities,
    static_lty = NULL,
    additional_controls = NULL,
    ylab = "fitted effect\nestimate (+ intercept) ±2*SE",
    ...) {
  if (surp_units == "bits") {
    data_to_plot <- data_to_plot %>% mutate(surprisal = surprisal / log(2))
    if (!is.null(controls)) {
      controls <- controls %>% mutate(surprisal = surprisal / log(2))
    }
    xlab <- "surprisal (bits)"; xbreaks <- seq(0, 50, 15)
  } else if (surp_units == "nats") {
    xlab <- "surprisal (nats)"; xbreaks <- seq(0, 30, 10)
  }
  if (use_shifted_est) data_to_plot <- mutate(data_to_plot, est = est_shifted)
  p <- data_to_plot %>%
    ggplot(aes(x = surprisal, y = trans(est),
               ymin = trans(est - se.mult*se), ymax = trans(est + se.mult*se),
               fill = context_len))
  if (!is.null(controls)) {
    # Add in linear data
    if (use_shifted_est) controls <- mutate(controls, est = est_shifted)
    p <- p +
      geom_ribbon(data = controls, aes(y=NULL), 
                  fill = linear_control_color, size = 0, alpha = 0.1) +
      geom_line(data = controls, 
                lty= linear_control_lty, color = linear_control_color, 
                size = linear_control_lsize, alpha= 1.0)
  }
  if (!is.null(additional_controls)) {
    # Add in other data
    if (use_shifted_est) additional_controls <- mutate(
      additional_controls, est = est_shifted)
    p <- p +
      geom_ribbon(data = additional_controls, aes(y=NULL), 
                  # fill = additional_control_color, 
                  size = 0, alpha = 0.1) +
      geom_line(data = additional_controls, aes(color = context_len),
                lty = additional_control_lty, #color = additional_control_color, 
                size = additional_control_lsize, alpha= 1.0)
  }
  if (!is.null(density_df)) {
    max_surps = data_to_plot %>% group_by(model, context_len) %>%
      summarise(max_surp=max(surprisal), .groups="drop")
    density_df <- density_df %>% 
      full_join(max_surps, by = c("model", "context_len")) %>%
      filter(surprisal <= max_surp)
    density_ymin <- if (use_shifted_est) 312 else -10
    density_df <- mutate(density_df, density = 120 * density + density_ymin)
    p <- p +
      geom_line(
        data = density_df %>% mutate(est_type = "mean"), 
        aes(ymin = NULL, ymax = NULL, y = density, x = surprisal, linetype = NULL), 
        linetype = "solid",
        color = density_color, alpha=.35)
    p <- p +
      geom_ribbon(data = density_df %>% mutate(est_type = "mean"),
                aes(x = surprisal, y = NULL, # forget the y aesthetic
                    ymin = density_ymin, ymax = density),
                fill = density_color, alpha = 0.2)
  }
  p <- p + geom_ribbon(aes(y=NULL), alpha = 0.2, size=0)
  p <- if (!is.null(static_lty)) {
        p + geom_line(aes(color=context_len), linetype = static_lty)
      } else {
        p + geom_line(aes(color=context_len, linetype = est_type))
        }
  p <- p +
    scale_fill_discrete(name = "context amount", labels = context_len_labs) +
    scale_color_discrete(name = "context amount", labels = context_len_labs) +
    scale_linetype_discrete(name = "effect", labels = est_type_labs) +
    facet_grid(..., labeller = my_labeller) +
    scale_x_continuous(breaks = xbreaks) +
    labs(x = xlab, y=ylab)
}

plot_easy_mean_patch <- function(
      data_to_plot, data_to_plot_control,
      order_models_by_PPL = TRUE,
      ctx_lens = context_lens, density_df = surprisal_densities, 
      meanbreaks = seq(-25, 500, 25), ...){
    if (order_models_by_PPL) {
      data_to_plot$model <- factor(
        data_to_plot$model, levels = models_in_desc_PPL_order)
      data_to_plot_control$model <- factor(data_to_plot_control$model, levels = models_in_desc_PPL_order)
    }
     
    which_descr <- "mean RT"
    data_to_plot <- data_to_plot %>% filter(est_type=="mean")
    data_to_plot_control <- data_to_plot_control %>% filter(est_type=="mean")
    space <- "free"; ybreaks <- meanbreaks
    static_lty = "solid"; ltyguide <- "none"
    data_to_plot %>% 
      filter(context_len %in% ctx_lens) %>%
      plot_data(controls = data_to_plot_control %>%
                  filter(context_len %in% ctx_lens),
                density_df = density_df %>% filter(context_len %in% ctx_lens),
                static_lty = static_lty,
                rows =
                  context_len  #+ est_type
                ~   model,
                scales="free", space=space,
                switch="y",
                ylab = "mean RT ±2*SE",
                ...) +
      labs(#title = "GAM fits of the effect of surprisal on reading time",
           subtitle = glue::glue("Partial effect of surprisal on {which_descr}")) +
      theme(
        strip.placement='outside',
        # axis.title.x = element_blank(), axis.text.x = element_blank(),
        legend.position = c(.08,.7),
        legend.background = element_rect(size = 0, color = "white")) +
        # guides(linetype = ltyguide) +
      scale_y_continuous(breaks = ybreaks, position = "right")
}

plot_easy_var_patch <- function(
      data_to_plot, data_to_plot_control,
      order_models_by_PPL = TRUE,
      ctx_lens = context_lens, density_df = surprisal_densities, ...){
    if (order_models_by_PPL) {
      data_to_plot$model <- factor(
        data_to_plot$model, levels = models_in_desc_PPL_order)
      data_to_plot_control$model <- factor(
        data_to_plot_control$model, levels = models_in_desc_PPL_order)
    }
    varbreaks <- function(x) {if (max(x)<5.3) seq(4,6,.2) else seq(3,7, .5)}
    which_descr <- "log standard deviation in RT"
    data_to_plot <- data_to_plot %>% filter(est_type=="var")
    data_to_plot_control <- data_to_plot_control %>% filter(est_type=="var")
    space <- "free"; ybreaks <- varbreaks
    static_lty = variance_lty; ltyguide <- "none"
    data_to_plot %>%
      filter(context_len %in% ctx_lens) %>%
      plot_data(controls = data_to_plot_control %>%
                  filter(context_len %in% ctx_lens),
                density_df = NULL,
                ylab = "log (sd RT) ±2*SE",
                static_lty = static_lty,
                rows =
                  context_len #+ est_type
                ~   model,
                scales="free", space=space,
                switch="y",
                # trans = \(x){exp(x)^2},
                ...) +
      labs(#title = "GAM fits of the effect of surprisal on reading time",
           subtitle = glue::glue("Partial effect of surprisal on {which_descr}")) +
      theme(
        strip.placement='outside',  strip.text.x = element_blank(),
        legend.position = c(.08,.7),
        legend.background = element_rect(size = 0, color = "white")) +
        # guides(linetype = ltyguide) +
      scale_y_continuous(breaks = ybreaks, position = "right")
}


plot_easy_both_patch <- function(
    data_to_plot, data_to_plot_control,
    order_models_by_PPL = TRUE,
    ctx_lens = context_lens, density_df = surprisal_densities, ...) {
  p_mean <- plot_easy_mean_patch(data_to_plot, data_to_plot_control,
      order_models_by_PPL = order_models_by_PPL,
      ctx_lens = ctx_lens, density_df = density_df, ...) + 
    theme(axis.title.x = element_blank(), axis.text.x = element_blank())
  p_var <- plot_easy_var_patch(data_to_plot, data_to_plot_control,
      order_models_by_PPL = order_models_by_PPL,
      ctx_lens = ctx_lens, density_df = density_df, ...)
  p_mean + p_var +
    patchwork::plot_layout(ncol = 1, nrow = 2, heights = c(4,3)) +
    patchwork::plot_annotation(
    title = "GAM fits of the effect of surprisal on reading time"
  )
}
```


```{r}
#| warning: false

density_color <- "grey"
linear_control_color <- "black"
linear_control_lty <- "13"
linear_control_lsize <- .25
variance_lty <- "31"

additional_control_color <- "grey"
additional_control_lty <- "33"
additional_control_lsize <- .25

plot_easy_mean_patch(data_to_plot_var0_gauss_wprev, data_to_plot_lin0var_wprev,
                     meanbreaks = seq(-25, 650, 50)) +
  labs(y="RT (ms) ±2*SE",
       title = "Effect of surprisal on mean RT",
       subtitle = "GAMs with constant variance assumption")
### FIG FOR PAPER (3 contexts)
ggsave(
  paste0(figsdir, "gauss0var", context_pref, "_wprev", "_wlinear0var_both_patch.pdf"),
  width = 10.5, height = ifelse(twocontexts, 3.5, 4.5))
```


```{r}
#| code-fold: true

additional_control_color <- 3
additional_control_lty <- "33"
additional_control_lsize <- .25

plot_easy_mean_patch_3 <- function(
      data_to_plot, data_to_plot_control, data_to_plot_control2,
      order_models_by_PPL = TRUE,
      ctx_lens = context_lens, density_df = surprisal_densities,
      meanbreaks = seq(-25, 500, 25), ...){
    if (order_models_by_PPL) {
      data_to_plot$model <- factor(
        data_to_plot$model, levels = models_in_desc_PPL_order)
      data_to_plot_control$model <- factor(
        data_to_plot_control$model, levels = models_in_desc_PPL_order)
    }

    which_descr <- "mean RT"
    data_to_plot <- data_to_plot %>% filter(est_type=="mean")
    data_to_plot_control <- data_to_plot_control %>% filter(est_type=="mean")
    data_to_plot_control2 <- data_to_plot_control2 %>% filter(est_type=="mean")
    space <- "free"; ybreaks <- meanbreaks
    static_lty = "solid"; ltyguide <- "none"
    data_to_plot %>%
      filter(context_len %in% ctx_lens) %>%
      plot_data(controls = data_to_plot_control %>%
                  filter(context_len %in% ctx_lens),
                additional_controls = data_to_plot_control2 %>%
                  filter(context_len %in% ctx_lens),
                density_df = density_df %>% filter(context_len %in% ctx_lens),
                static_lty = static_lty,
                rows =
                  context_len  #+ est_type
                ~   model,
                scales="free", space=space,
                switch="y",
                ylab = "mean RT ±2*SE",
                ...) +
      labs(#title = "GAM fits of the effect of surprisal on reading time",
           subtitle = glue::glue("Partial effect of surprisal on {which_descr}")) +
      theme(
        strip.placement='outside',
        # axis.title.x = element_blank(), axis.text.x = element_blank(),
        legend.position = c(.08,.7),
        legend.background = element_rect(size = 0, color = "white")) +
        scale_y_continuous(breaks = ybreaks, position = "right") +
      guides(linetype = ltyguide)
}
```


```{r}
plot_easy_mean_patch_3(
  data_to_plot_var0_gauss_w3prev,
  data_to_plot_lin0var_wprev,
  data_to_plot_var0_gauss_wprev %>%  filter(est_shifted <= 500),
                     meanbreaks = seq(-25, 650, 50)) +
  labs(y="RT (ms) ±2*SE",
       title = "Effect of surprisal on mean RT, GAMs with constant variance",
       subtitle = "comparing using 3 words for spillover (solid) words to just 1 (dashed)")
### FIG FOR PAPER (3 contexts)
ggsave(
  paste0(figsdir, "gauss0var", context_pref, "_wprev_vs_wprev3_lin0var.pdf"),
  width = 10.5, height = ifelse(twocontexts, 3.5, 4.5))
```

Main plot figure for paper:

```{r}
plot_easy_both_patch(data_to_plot_smooth_wprevm, data_to_plot_linear_wprevm)
### FIG FOR PAPER (2 and 3 contexts)
ggsave(
  paste0(figsdir, "gaulss", context_pref, "_wprevm", "_wlinear_both_patch.pdf"),
  width = 10.5, height = ifelse(twocontexts, 6, 7.5))
```



## Superlinearity

```{r}
#| code-fold: true

make_half_linears_dataset <- function(data_to_plot, breakpoint=0.5, se.mult=1.96) {
  #" fit simple linear models to the smooth fits before and after the breakpoint
  #" breakpoint should be between 0 and 1 (proportion of max surprisal to break at)
  data_to_plot %>% 
    group_by(model, context_len) %>% 
    mutate(secondhalf = surprisal > max(surprisal) * breakpoint) %>%
    group_by(secondhalf, est_type, .add = TRUE) %>%
    do(f = broom::tidy(lm(est ~ surprisal, weights = 1/sqrt(se), data = .))) %>%
    tidyr::unnest(f) %>%
    mutate(lower = estimate - se.mult * std.error,
           upper = estimate + se.mult * std.error) %>%
    filter(term == "surprisal")  
}

# half_linears3 <- make_half_linears_dataset(data_to_plot_smooth_wprevm, breakpoint=1/3)
half_linears2 <- make_half_linears_dataset(data_to_plot_smooth_wprevm, breakpoint=1/2)

# For the datasubset_lessN versions (just GPT3d for now)
half_linears2_less6 <- make_half_linears_dataset(
  data_to_plot_datasubsetless6_smooth_wprevm, breakpoint=1/2)
half_linears2_less12 <- make_half_linears_dataset(
  data_to_plot_datasubsetless12_smooth_wprevm, breakpoint=1/2)

make_halflinears_diffs_dataset <- function(half_linears_dataset) {
  half_linears_dataset %>% group_by(model, context_len, est_type) %>%
    do(diff_slope_hi_v_low = 
         filter(., secondhalf==TRUE)$estimate - filter(., secondhalf==FALSE)$estimate,
       sum_std.error = 
         filter(., secondhalf==TRUE)$std.error + filter(., secondhalf==FALSE)$std.error
       ) %>%
    tidyr::unnest(c(diff_slope_hi_v_low, sum_std.error)) %>%
    mutate(upper = diff_slope_hi_v_low + 1.96 * sum_std.error,
           lower = diff_slope_hi_v_low - 1.96 * sum_std.error)
}

halflinear_diffs2 <- make_halflinears_diffs_dataset(half_linears2)
# halflinear_diffs3 <- make_halflinears_diffs_dataset(half_linears3)

## Save halflinear data
# halflinear_diffs2 %>% readr::write_csv("natural-stories-surprisals/gaulss_halflinear_diffs2.csv")
# halflinear_diffs3 %>% readr::write_csv("natural-stories-surprisals/gaulss_halflinear_diffs3.csv")
```


```{r}
#| code-fold: true

## Read in halflinear data
# halflinear_diffs2 <- load_in_data("natural-stories-surprisals/gaulss_halflinear_diffs2.csv")
# halflinear_diffs3 <- load_in_data("natural-stories-surprisals/gaulss_halflinear_diffs3.csv")

halflinear_diffs2 %>% filter(context_len %in% context_lens) %>% 
  ggplot(aes(x = model, y = diff_slope_hi_v_low, shape = est_type,
             ymin = lower, ymax = upper, color = context_len)) +
  geom_hline(yintercept=0, color="gray") +
  geom_errorbar(width = .25) + geom_point(size=3, alpha = .4) +
  facet_grid(context_len +est_type~ model, labeller = my_labeller, scales = "free") +
  scale_fill_discrete(name = "context amount", labels = context_len_labs) +
  scale_color_discrete(name = "context amount", labels = context_len_labs) +
  scale_shape_discrete(name = "", labels = est_type_labs) +
  theme(
    legend.position = c(0.08, 0.7),
    axis.text.x = element_blank(),
    legend.background = element_rect(size = 0, color = "white")) +
  labs(
    x = "model",
    y = "slope at high surprisal - slope at low surprisal",
    title = "Superlinearity by LM",
    subtitle = "Difference in slope of smooth predictions for high vs. low surprisal")
# ggsave(
#   paste0(
#     figsdir, "", context_pref, "gaulss_halflinear_diffs3.pdf"),
#   width = 5, height = 4)
```

#### Cutoffs most surprising words

```{r}
#| code-fold: true
#| warning: false

data_to_plot_multicutoffs <- bind_rows(
  data_to_plot_smooth_wprevm, # cutoff=NA for these rows
  data_to_plot_datasubsetless6_smooth_wprevm,
  data_to_plot_datasubsetless12_smooth_wprevm
  ) %>% 
  mutate(cutoff = forcats::fct_rev(forcats::as_factor(cutoff)))

compare_one_model_multi_cutoffs_surprisal<-function(
    data_to_plot_multicutoffs, 
    use_shifted_est = T, trans = I, se.mult = 1.96,
    ctx_lens = context_lens, density_df = surprisal_densities, 
    ylab = "fitted effect\nestimate (+ intercept) ±2*SE",...) {
  
  d <- data_to_plot_multicutoffs %>% 
    filter(model=="GPT3-davinci", context_len=="max", est_type == "mean")
  l<- levels(d$cutoff) %>% as.numeric()
  if (use_shifted_est) d <- mutate(d, est = est_shifted)
  p <- d %>%
    ggplot(aes(x = surprisal, y = trans(est),
               color=cutoff, #linetype = !is.na(cutoff),
               ymin = trans(est - se.mult*se), ymax = trans(est + se.mult*se)))
  
  max_surps <- d %>% 
    group_by(model, context_len, cutoff) %>% 
    summarise(max_surp=max(surprisal), .groups="drop")
  density_df <- density_df %>% 
    right_join(max_surps, by = c("model", "context_len")) %>% 
    filter(surprisal <= max_surp)
  density_ymin <- if (use_shifted_est) 312 else -10
  density_df <- mutate(density_df, density = 120 * density + density_ymin)
  
  p <- p +
    geom_line(data = density_df %>% mutate(est_type = "mean"),
              aes(ymin = NULL, ymax = NULL, color = cutoff,
                  y = density, x = surprisal),
              linetype = "solid", 
              alpha=.25
              )
  p <- p +
    geom_ribbon(data = density_df %>% mutate(est_type = "mean"), inherit.aes = F,
              aes(x = surprisal, 
                  lty = NULL,
                  ymin = density_ymin, ymax = density),
              fill = density_color, size = 2,
              alpha = 0.35 / length(l))
  
  p <- p + 
    # ggplot2::annotate(
    #   geom="text", x=l+0.1, y=312+10, 
    #   label=glue::glue("surprisal cutoff = {l}"), 
    #   color= "grey", size=3, vjust=1, hjust=0) +
    ggplot2::annotate(
      geom="segment", x=l, xend=l, y=-Inf, yend=322, color="grey")
  
  p +
    geom_line(alpha = 0.5) +
    geom_ribbon(aes(y=NULL, color = NULL, fill = cutoff), alpha = 0.2, size = 0) +
    # scale_fill_discrete(name = "cutoff",  breaks = c(NA, l), labels = c("none", l)) +
    # scale_color_discrete(name = "cutoff", breaks = c(NA, l), labels = c("none", l)) +
    scale_fill_manual(name = "cutoff", values=rev(scales::hue_pal()(14)), 
                   breaks = c(5:18,NA), 
                   labels = c(5:18,"none")) +
    scale_color_manual(name = "cutoff", values=rev(scales::hue_pal()(14)), 
                   breaks = c(5:18,NA), 
                   labels = c(5:18,"none")) +
    scale_linetype_manual(name = "whether cutoff", values = c(`TRUE`=1, `FALSE`=3)) +
    guides(linetype = "none", size = "none") +
    labs(x = "surprisal (nats)", y = ylab)
}
p_gpt3_cutoffs_meaneffect <- compare_one_model_multi_cutoffs_surprisal(data_to_plot_multicutoffs) +
  facet_grid( ~ model, labeller = my_labeller, scales = "free_y") +
    theme(
    legend.position = c(.24,.68),
    legend.background = element_rect(size = 0, fill = alpha("white", 0.5), color = alpha("white", 0.5))
    ) +
  labs(
    y = "effect on mean RT ±2*SE",
    subtitle="Effect of surprisal on RT", 
    # title="GAMs refit without highest surprisal items"
    )
# ggsave(p_gpt3_cutoffs_meaneffect,
#   filename = paste0(figsdir, "gpt3d_surprisal_cutoffs_effect_on_mean", ".pdf"),
#   width = 3, height = 3.5)
```


### Superlinearity vs perplexity

```{r}
#| code-fold: true
#| warning: false

d <- full_join(halflinear_diffs2, surprisal_means, by = c("model", "context_len")) %>% 
  filter(context_len %in% context_lens, est_type == "mean") %>% 
  rename(diff = diff_slope_hi_v_low, sem = sum_std.error)
d %>% 
  filter(model %>% str_starts("GPT")) %>% 
  #  note: `weight` aesthetic will be passed on to lm's `weights`` in geom_smooth/geom_line
  ggplot(aes(x = -mean_surprisal, y = diff, color = context_len,
             ymax = upper, ymin = lower, shape = context_len,
             weight= 1/sqrt(sem))) +
  # facet_grid(est_type ~ ., labeller = my_labeller, scales = "free_y") +
  # ggpubr::stat_regline_equation(label.x.npc = .8, label.y.npc = .8, 
  #                               aes(label = after_stat(eq.label))) +
  ggpubr::stat_regline_equation(
    label.x.npc = .6, label.y.npc = .95, 
    aes(color=NULL, shape =NULL, label = after_stat(rr.label)), 
    color="grey", size=4, show.legend  = FALSE) +
  geom_smooth(
    aes(color=NULL,shape =NULL), 
    method = "lm", formula = 'y~x', size = 0, alpha = .1, color="#f1f1f1") + # SE ribbon, no line
  geom_line(
    aes(color=NULL, shape =NULL), stat = "smooth", 
    method = "lm", formula = 'y~x', alpha = .66, lty="dotted", color="grey") + # just line
  geom_errorbar(data = d, width = .03) + 
  geom_point(data = d, aes(fill = str_starts(model,"GPT")), size=3, alpha = .6) +
  ggrepel::geom_text_repel(data = d,
    aes(label = model_labs[model]),
    max.overlaps = Inf, min.segment.length = 0.2, size = 3, segment.alpha = .2,
    force_pull = .5, hjust = .9,  box.padding = .5, color = "black", alpha = 0.4) +
  theme(legend.position="top") +
  scale_color_discrete(name = "context amount", labels = context_len_labs) +
  scale_fill_manual(name = "GPT", values = c(`TRUE`="grey", `FALSE`="white")) +
  scale_shape_manual(
    values=if (twocontexts) c(21, 23) else c(21,22,23), 
    name = "context amount", labels = context_len_labs) +
  labs(
    x = "LM quality (negative log perplexity)",
    y = "superlinearity",
    title = "Superlinearity vs. LM quality") +
  guides(fill = "none")
```


```{r}
#| warning: false
### FIG FOR PAPER
ggsave(
  paste0(
    figsdir, context_pref,
    "gaulss_halflinear_diffs2_justFitGPTs_mean_v_mean_surprisal_lengthstogether.pdf"),
 width = 7, height = 4.25)
```

#### Cutoffs superlinearity

```{r}
#| code-fold: true

halflinear_diffs2_less6  <- make_halflinears_diffs_dataset(half_linears2_less6)
halflinear_diffs2_less12 <- make_halflinears_diffs_dataset(half_linears2_less12)

p_gpt3d_cutoffs_superlinearity <- bind_rows(
  halflinear_diffs2 %>% mutate(cutoff = NA) %>% 
    filter(model == "GPT3-davinci", context_len == "max"),
  halflinear_diffs2_less6 %>% mutate(cutoff = 6),
  halflinear_diffs2_less12 %>% mutate(cutoff = 12)
  ) %>% 
  filter(est_type == "mean") %>% 
  mutate(cutoff = factor(cutoff)) %>% 
  filter(context_len %in% context_lens) %>% 
  rename(diff = diff_slope_hi_v_low, sem = sum_std.error) %>%
  ggplot(aes(x = cutoff, y = diff, color = cutoff, 
             ymax = upper, ymin = lower)) +
  geom_errorbar(width = .03) + geom_point(size=3, alpha = .6) +
  facet_grid( ~ model, labeller = my_labeller, scales = "free_y") +
  scale_color_manual(name = "cutoff", values=rev(scales::hue_pal()(14)), 
                     breaks = c(5:18,NA), 
                     labels = c(5:18,"none")) +
  scale_x_discrete(name = "cutoff", 
                   breaks = c(5:18,NA), 
                   labels = c(5:18,"none")) +
  expand_limits(y=c(0, NA)) +
  theme(legend.position = "none") + 
  labs(
    # x = "Cutoff",
    y = "superlinearity\nof effect on mean RT",
    subtitle = "Superlinearity"
    # title = "GAMs refit without highest surprisal items"
    )
# 
# ggsave(p_gpt3d_cutoffs_superlinearity,
#   filename = paste0(figsdir, "gpt3d_surprisal_cutoffs", ".pdf"),
#   width = 3, height = 3.5)
wrap_plots(
  p_gpt3_cutoffs_meaneffect, 
  p_gpt3d_cutoffs_superlinearity, guides = "collect", widths = c(1.2,1)) + 
  plot_annotation(
    title="GAMs refit without highest surprisal items")
### FIG FOR PAPER
ggsave(
  filename = paste0(figsdir, "gpt3d_surprisal_cutoffs", ".pdf"),
  width = 6, height = 3.5)
```


## Linear model coefficients

```{r}
#| warning: false
summaries_linear <- readr::read_rds(
  paste0(datadir, "model-summaries-NEW-linear-wprevm.rds"))

linear_coefs<-purrr::map_df(
     summaries_linear, 
     \(s) {s$p.table %>% as_tibble() %>% mutate(term=names(s$p.pv))},
     .id = "model")
linear_coefs %>%
  separate_modelnames() %>%
  filter(context_len %in% context_lens) %>% 
  ggplot(aes(x=model,y=Estimate,#y=`Pr(>|z|)`,
             color=context_len)) +
  geom_point() + facet_wrap(~term, scales = "free", labeller=my_labeller) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_color_discrete(name = "context amount", labels = context_len_labs) 

# Whether to include both mean and variance
which_terms <- "both" # specify either "surprisal", "surprisal.1", or "both"
terms_included <- if (which_terms=="both") c("surprisal", "surprisal.1") else which_terms

linear_coefs %>% 
  separate_modelnames() %>%
  mutate(model = dplyr::recode(model, !!!model_labs)) %>% 
  filter(context_len %in% context_lens) %>% 
  filter(term %in% terms_included) %>% 
  ggplot(aes(x=model,y=Estimate,
             ymin=Estimate-1.96*`Std. Error`,
             ymax=Estimate+1.96*`Std. Error`,
             color=context_len)) +
  geom_point(aes(shape=context_len)) + 
  geom_errorbar(width = .25, alpha=0.75) + 
  facet_grid(term~., scales = "free", labeller=my_labeller) +
  scale_color_discrete(name = "context amount", labels = context_len_labs) +
  scale_shape_manual(
    values=if (twocontexts) c(1, 5) else c(1,0,5), name = "context amount", 
    labels = context_len_labs) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position="top")  +
  labs(x = "LM", y = "coefficient estimate",
       title = paste0("Effect of surprisal",
                      " on RT and variance in RT"), #<- hardcoded for which_terms="both"
       subtitle="from linear control GAMs")
### FIG FOR PAPER (2 contexts)
ggsave(
  paste0(figsdir, context_pref, "gaulss_linear_coefs_", which_terms, ".pdf"),
  width = 5, height = if (which_terms == "both") 4.5 else 3)
```

## Autocorrelation plots

```{r}
#| message: false
# Load a model to inspect (P)ACF of
m_wprevmk6_gpt3d <- readr::read_rds(
  paste0(datadir, "GAM_models/GPT3-davinci.rds"))

# Get the data on which the model was trained:
fitdata_perword <- readr::read_rds(
  paste0(datadir, "surps_lms.RTs.perword.rds"))
fitdata <- readr::read_rds(
  paste0(datadir, "surps_lms.RTs.rds"))

# file prepare_data.R defines functions
# - `prepare_data_with_perword`
# - `sort_data_for_tokenization_type`
source("utils/prepare_data.R")

# prepare data as it was prepared for model when trained
data_prepared_gpt3d <- prepare_data_with_perword(
  data = fitdata, data.perword = fitdata_perword, 
  modelname = "GPT3-davinci", how_many_prevs = 1)
```


Make function to (gg)plot ACF
```{r}
#| code-fold: true

get_clim <- function(x, ci=0.95, ci.type="white"){
  #' Gets confidence limit data from acf object `x`
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  if (class(x) != "acf") stop('pass in object of class "acf"')
  clim0 <- qnorm((1 + ci)/2) / sqrt(x$n.used)
  if (ci.type == "ma") {
    clim <- clim0 * sqrt(cumsum(c(1, 2 * x$acf[-1]^2))) 
    return(clim[-length(clim)])
  } else {
    return(clim0)
  }
}

ggplot_acf <- function(
    x,
    ci=0.95, ci.type="white", ci.col = "blue"){
  #' Replicates plot.acf() but using ggplot by default instead of base R plot
  #' `x` must be an object of class "acf" such as that outputted by `acf()`
  #' `ci.type` must be "white" or "ma"
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  if (class(x) != "acf") stop('pass in object of class "acf"')
  with.ci <- ci > 0 && x$type != "covariance"
  with.ci.ma <- with.ci && ci.type == "ma" && x$type == "correlation"
  if(with.ci.ma && x$lag[1L, 1L, 1L] != 0L) {
    warning("can use ci.type=\"ma\" only if first lag is 0")
    with.ci.ma <- FALSE
  }
  clim <- get_clim(x, ci=ci, ci.type=ci.type)
  df <- data.frame(lag = x$lag, acf=x$acf)
  p <- ggplot(df, aes(x=lag)) +
    geom_linerange(aes(ymax=acf, ymin=0)) +
    labs(y="ACF", x="Lag")
  if (with.ci) {
    if (ci.type == "white") {
      p <- p + 
        geom_hline(yintercept = 0-clim, lty = 2, col = ci.col) +
        geom_hline(yintercept = 0+clim, lty = 2, col = ci.col)
    } else if (with.ci.ma && ci.type == "ma") { # ci.type="ma" not allowed for pacf
      dfclim <- df[-1,]
      dfclim$clim <- clim
      p <- p + 
        geom_line(data = dfclim, aes(y = 0-clim), lty = 2, col = ci.col) +
        geom_line(data = dfclim, aes(y = 0+clim), lty = 2, col = ci.col)
    }
  }
  return(p)
}
```


```{r acf_resid_gg}
acf_resid_gg <- function(
    x, split_pred, dat, plot_sd=F, sig=F, series_length=NULL, ...){
  #' x should be either a model, that is: class(x) = "bam" "gam" "glm" "lm" 
  #'             or a numeric vector (of residuals, say)
  #' sig is the significance upper and lower lines (+- 1.96)
  #' series_length is the length of the series needed for sig
  if(is(x,"lm")){
    x <- resid(x)
  } else {
    if (!is(x,"numeric")) {
      stop(glue::glue("Don't know what to do with input object of class {class(x)}"))
    }
  }
  if(is.null(dat))
    dat <-tibble(x=x,response=NaN)
  dat$x <- NA
  dat[!is.na(dat$response),]$x <- x
  acf <- itsadug::acf_plot(dat$x, plot=F, split_by = split_pred, ...)
  df <- tibble(acf=acf,lag=seq(0,length(acf)-1))
  p <- df %>% 
    ggplot(aes(x=lag,y=acf)) +
    geom_col(width=0.1, alpha=0.9) +
    geom_point()
  if(plot_sd){
    df <- df %>% 
      mutate(sd=itsadug::acf_plot(dat$x, fun=sd, plot=F, split_by = split_pred, ...))
    p <- df %>% 
    ggplot(aes(x=lag,y=acf)) +
      geom_col(width=0.4, alpha=0.5) + 
      geom_errorbar(aes(ymin=acf-1.96*sd,ymax=acf+1.96*sd), width=0.3, color="red") 
  }
  if(sig){
    if(is.null(series_length)) { stop(glue::glue("You must specify series length")) }
    sig <- qnorm(1 - 0.05 / 2) / sqrt(series_length)
    p <- p +
      geom_hline(yintercept = 0 - sig, lty = 2, color = "blue") +
      geom_hline(yintercept = 0 + sig, lty = 2, color = "blue")
  }
  return(p)
}
```

Data ordering
: When plotting GAM fits (with `plot.gam()` for example) it doesn't matter what order the data is in but when plotting ACF of residuals the order of the data does matter. So be careful (note, for `itsadug::acf_resid`, with the right `split_pred` selected, it will work).

```{r add_resids}
# Let's get the residuals and then put them in the right order,
# that is, sort by subject > story_num > word_num_in_story.

assertthat::are_equal( # note: already removed NAs so nothing to worry about there
  data_prepared_gpt3d %>% nrow,
  resid(m_wprevmk6_gpt3d) %>% length) 

data_prepared_gpt3d_withresids <- data_prepared_gpt3d
data_prepared_gpt3d_withresids$resids <- resid(m_wprevmk6_gpt3d)
data_resorted <- arrange(
  data_prepared_gpt3d_withresids, subject, story_num, word_num_in_story)
```

Thee dataframe `data_resorted` will now work for getting ACF of resids.  Just one more thing: We just need to subset the data to remove the handful of observations from workers who only started stories, but didn't get far (this messes up getting ACF plots for reasons I'm unsure of). Since it's only a very small number of items, this won't affect the autocorrelation results much.

```{r subset_workers}
subset_workers<-function(nItems_threshold_productive = 1000, data=fitdata){
  how_many_stories_per_worker <- data %>% group_by(WorkerId) %>%
    summarise(n_stories=n_distinct(story_num)) %>% arrange(desc(n_stories))
  how_many_items_per_worker <- data %>% group_by(WorkerId) %>%
    count() %>% arrange(desc(n)) %>%
    mutate(is_productive = n > nItems_threshold_productive)
  workers_info <- full_join(
    how_many_stories_per_worker, how_many_items_per_worker, by = "WorkerId") %>%
    arrange(desc(is_productive), desc(n_stories), 
            (n-median(how_many_items_per_worker$n))^2)
  return(workers_info)
}

workers_info <- subset_workers(nItems_threshold_productive = 1500)
excluded_workers_info <- workers_info %>% filter(!is_productive | n_stories <= 2)
selected_workers_info <- workers_info %>% filter(is_productive, n_stories > 2)
cat("Excluding workers with few or sparse observations,",
    "for calculating ACF.\n proportion of workers excluded = ", 
    length(excluded_workers_info$n),"out of", length(workers_info$n),"\n",
    "proportion of items excluded = ", 
    sum(excluded_workers_info$n)/sum(workers_info$n),"\n")
knitr::kable(selected_workers_info)

# Unsorted
data_prepared_gpt3d_withresids_sub <- data_prepared_gpt3d_withresids %>%
  filter(WorkerId %in% selected_workers_info$WorkerId)
# Sorted
data_resorted_subset <- data_resorted %>% 
  filter(WorkerId %in% selected_workers_info$WorkerId)
```

Example comparing the necessity of sorting

```{r plot_example}
#| code-fold: true
data_prepared_gpt3d_withresids %>% 
  pull(resids) %>% 
  itsadug::acf_plot(
    split_by = list(
      data_prepared_gpt3d_withresids$WorkerId, 
      data_prepared_gpt3d_withresids$story_num),
    max_lag = 17,
    main = "ACF of GPT3-davinci GAM resids without necessary reodering"
  )
data_resorted_subset <- data_resorted %>% 
  filter(WorkerId %in% selected_workers_info$WorkerId)
data_resorted_subset %>% 
  pull(resids) %>% 
  itsadug::acf_plot(
    split_by = list(
      data_resorted_subset$WorkerId, 
      data_resorted_subset$story_num),
    max_lag = 17,
    main = "ACF of GPT3-davinci GAM resids - correctly reordered"
  )
```

A partial autocorrelation function plotter
: In `my_pacf_plot.R` I modify the code from `itsadug::acf_plot()` to calculate PACF instead of ACF. That is, it will get mean `pacf` across splits instead of getting complete `acf`. 

```{r my_pacf_plot}
# defines function `my_pacf_plot`
source("utils/my_pacf_plot.R")
```

```{r plot_avg_ACF}
plot_avg_ACF <- function(all_ACF, with.sem = F, ylim=c(NA, 1), xlim=c(0, NA)){
  # For plotting aggregated [p]acf with input `all_ACF <- plot_[p]acf( _ , return_all=T)`
  # mean with sem bars.
  ci <- 0.95
  with.ci <- T
  ci.col <- 'blue'

  plot_df <- all_ACF$dataframe %>% 
    group_by(lag) %>% 
    summarise(
      mean = mean(acf),
      sd = sd(acf),
      n = n(),
      sem = sd / sqrt(n),
      sem.upper = mean + qnorm((1 + 0.95)/2) * sem,
      sem.lower = mean - qnorm((1 + 0.95)/2) * sem,
      mean_ci = mean(ci), # with default ci.type= "white" these are all the same value
      .groups = "drop") %>%
    drop_na(mean)
  clim <- plot_df$mean_ci[[1]]
  p <- plot_df %>%
    ggplot(aes(x = lag, y = mean)) +
    geom_linerange(aes(ymax = mean, ymin=0)) +
    geom_hline(yintercept = 0) + 
    lims(y = as.numeric(ylim),
         x = as.numeric(xlim))
  if (with.ci) {
    p <- p +
        geom_hline(yintercept = 0-clim, lty = 2, col = ci.col) +
        geom_hline(yintercept = 0+clim, lty = 2, col = ci.col)
  }
  if (with.sem) {
    p <- p +
      geom_linerange(
        aes(ymin = sem.lower, ymax = sem.upper), 
        linewidth = 3, alpha=0.5, color = "red") 
  }
return(p)
}
```

Comparison of plotting functions for sanity
```{r}
data_resorted_subset_resids_ACF <- itsadug::acf_plot(
  data_resorted_subset$resids,
  split_by = list(
    data_resorted_subset$subject,
    data_resorted_subset$story_num),
  plot = T,
  return_all = T)
data_resorted_subset_resids_PACF <- my_pacf_plot(
  data_resorted_subset$resids,
  split_by = list(
    data_resorted_subset$subject,
    data_resorted_subset$story_num),
  plot = T,
  return_all = T, main="Partial ACF")

# For comparison
plot_avg_ACF(data_resorted_subset_resids_ACF) + 
  labs(
    title = "Mean ACF of residuals across stories, and subjects",
    subtitle = "GPT3-davinci",
    x = "Lag", 
    y="mean ACF")
plot_avg_ACF(data_resorted_subset_resids_PACF) + 
  labs(
    title = "Mean PACF of residuals across stories, and subjects",
    subtitle = "GPT3-davinci",
    x = "Lag", 
    y="mean Partial ACF")
```

```{r}
# Plot ACF mean and 95%CI
plot_avg_ACF(data_resorted_subset_resids_ACF, with.sem = T) + 
  labs(
    title = "ACF of residuals",
    subtitle = "Mean across stories, subjects; GPT3-davinci",
    x = "Lag",
    y="mean ACF")
### FIG FOR PAPER
ggsave(
  filename = paste0(figsdir, "gpt3d_ACF", ".pdf"),
  width = 4, height = 3)
plot_avg_ACF(data_resorted_subset_resids_PACF, with.sem = T) + 
  labs(
    title = "PACF of residuals",
    subtitle = "Mean across stories, subjects;  GPT3-davinci",
    x = "Lag", 
    y="mean Partial ACF")
### FIG FOR PAPER
ggsave(
  filename = paste0(figsdir, "gpt3d_PACF", ".pdf"),
  width = 4, height = 3)
plot_avg_ACF(data_resorted_subset_resids_PACF, with.sem = T, ylim=c(NA,NA)) + 
  labs(
    title = "PACF of residuals",
    subtitle = "Mean across stories, subjects;  GPT3-davinci",
    x = "Lag", 
    y="mean Partial ACF")
# ggsave(
#   filename = paste0(figsdir, "gpt3d_PACF_zoomed", ".pdf"),
#   width = 4, height = 3)
```


# Schematic diagram


Make fake data for illustration, where true relationship is superlinear

```{r}
#| code-fold: true

myarrow <- arrow(10, unit(4,"mm"), type= "closed")

# wrapper for stats::rbinom to sample logical bernoulli random variates
# should be identical to deprecated purrr::rbernoulli
rbern <- function(n, p = 0.5) {
  as.logical(rbinom(n, size = 1, prob = p))
}

gen_data <- function(N=5000, seed = 4){
  set.seed(seed)
  b0 <- .9; b1 <- -.11;
  lnk <- "inverse"
  if (lnk == "identity") {
    invlnk <- function(eta) eta
    lnk_fn <- function(mu) mu
    } else if (lnk == "inverse") {
    invlnk <- function(eta) 1 / eta
    lnk_fn <- function(mu) 1 / mu
    } else if (lnk == "log") {
    invlnk <- function(eta) exp(eta)
    lnk_fn <- function(mu) log(mu)
  }
  shape <- 10
  x<-rlnorm(N,meanlog=-1,sdlog=0.8)
  # x<-runif(N, 0,8)
  # x<-seq(0,8,8/(N-1))
  y_true = invlnk(b0 + b1 * x) # function of the linear expression
  # y_true = 1/(b0+b1*x)
  df<-data.frame(
    x,
    y_true,
    y.1 = 300 + 100 * rgamma(N, rate = 1.30 * shape / y_true, shape = shape),
    y.2 = 300 + 100 * rgamma(N, rate = 1.70 * shape / y_true, shape = shape),
    y.3 = 300 + 100 * rgamma(N, rate = 1.08 * shape / y_true, shape = shape),
    y.4 = 300 + 100 * rgamma(N, rate = 1.09 * shape / y_true, shape = shape),
    y.5 = 300 + 100 * rgamma(N, rate = 1.10 * shape / y_true, shape = shape)
  ) %>%
    pivot_longer(
      starts_with("y."), 
      names_to = "subj", values_to = "y_obs", names_prefix = "y.") %>%
    mutate(subj=factor(subj))
  # rm(x,y_true)
  return(df)
}
df <- gen_data()
ggplot(df) + geom_point(aes(x,y_obs,color=subj), alpha=.1) +
  geom_point(aes(x=x,y=300+100*y_true),shape=1,size=2)

```

```{r}
#| echo: false
#| eval: false
# Unused. Less realistic version, where x values are modified randomly per subj

df_10000 <- gen_data(N = 10000, seed = 4)

set.seed(4)
df_modified_NOTBYSUBJ <- df_10000 %>%
  mutate(selected=rbern(nrow(df_10000), p=0.002),
         x0 = x,
         x = if_else(selected, 
                     x + runif(nrow(df_10000), min=x/4,max=5.5-x),
                     x))

df_modified_NOTBYSUBJ %>% 
  ggplot(aes(x=x0,y=x0,alpha=selected, shape=selected)) + 
  geom_segment(data=df_modified_NOTBYSUBJ %>% filter(selected==TRUE), 
               aes(yend=x0, xend=x), 
               arrow = myarrow,
               alpha=0.5, color="red") +
  geom_point() + geom_point(aes(x=x))

df_modified_NOTBYSUBJ %>% 
  ggplot(aes(x=x0,y=y_obs,alpha=selected, color=subj)) + 
  geom_segment(data=df_modified_NOTBYSUBJ %>% filter(selected==TRUE), 
               aes(yend=y_obs, xend=x),
               arrow = myarrow,
               alpha=0.5) +
  geom_point() + geom_point(aes(x=x))
```
Modify to (mostly) overestimate some surprisals


```{r}
#| warning: false
set.seed(4)
df_modified <- df %>% select(-y_true) %>%
  pivot_wider(names_from=subj, values_from=y_obs) %>% 
  mutate(selected=rbern(nrow(.), p=0.004),
         x0 = x,
         x = if_else(selected, max(x0)*rbeta(nrow(.),1,1),x0)) %>% 
  pivot_longer(cols = c(`1`,`2`,`3`,`4`,`5`), names_to = "subj", values_to = "y_obs") %>% 
  mutate(subj = factor(subj)) %>% 
  mutate(across(where(~!is.numeric(.)),factor))

df_modified_longer <- df_modified %>% 
  pivot_wider(names_from=subj, values_from=y_obs) 
```

::: {.panel-tabset}
### Modified surprisals
```{r}
#| warning: false
#| code-fold: true
df_modified_longer %>% 
  ggplot(aes(x=x0,y=x0,alpha=selected,shape=selected)) + 
  geom_segment(data=df_modified_longer %>% filter(selected==TRUE), 
               aes(yend=x0, xend=x), 
               arrow = myarrow,
               alpha=0.5, color="red") +
  geom_point() + geom_point(aes(x=x))

```

### Plotted against response
```{r}
#| warning: false
#| code-fold: true
df_modified %>% 
  ggplot(aes(x=x0,y=y_obs,alpha=selected, color=subj)) + 
  geom_segment(data=df_modified %>% filter(selected==TRUE), 
               aes(yend=y_obs, xend=x),
               arrow = myarrow,
               alpha=0.5) +
  geom_point() + geom_point(aes(x=x))
```
:::

### Fit GAMs

```{r}
#| warning: false
# GAM formula
f <- y_obs ~  s(x, k=6) + s(x, subj, bs="fs", m=1)

m  <- mgcv::gam(formula = f, data=df)
m_ <- mgcv::gam(formula = f, data=df_modified)
```

### Plot

```{r plot}
#| warning: false
#| code-fold: true

## Could plot with gratia
# patchwork::wrap_plots(
#   gratia::draw(m, residuals = F, select = 1, rug=F),
#   gratia::draw(m_, residuals = F, select = 1, rug=F), nrow = 2)

## But just get ests out instead, so we can plot them together
ests <- gratia::smooth_estimates(m) %>% 
  filter(smooth=="s(x)") %>% mutate(est_shifted = est + coef(m)["(Intercept)"])
ests_ <- gratia::smooth_estimates(m_) %>% 
  filter(smooth=="s(x)") %>% mutate(est_shifted = est + coef(m)["(Intercept)"])

set.seed(4)
col <- c("red", "darkblue", "darkred")
ggplot() + 
  geom_hex(data = df_modified, aes(x,y=y_obs), size=0, bins=200, alpha=0.8) +
  scale_fill_gradient(low="#DDDDDD",high="black",guide = "none")+
  geom_line(data = ests, aes(x=x, y=est_shifted),color=col[2]) + 
  geom_ribbon(
    data = ests, 
    aes(x=x, y=est_shifted, ymin=est_shifted-se*1.96, ymax=est_shifted+se*1.96), 
    alpha=0.2, fill=col[2]) +
  geom_line(
    data = ests_, aes(x=x, y=est_shifted), color = col[1]) + 
  geom_ribbon(
    data = ests_, 
    aes(x=x, ymin=est_shifted-se*1.96, ymax=est_shifted+se*1.96), 
    alpha=0.2, fill=col[1]) +
  geom_point(
    data = df_modified %>% filter(selected==T), 
    aes(x=x,y=y_obs),color=col[1], alpha = 0.1) +
  geom_point(
    data = df_modified %>% filter(selected==T), 
    aes(x=x0,y=y_obs),color=col[2], alpha = 0.1) +
  geom_segment(
    data = df_modified %>% filter(selected==T), 
    aes(x = x0, xend=x,y=y_obs,yend=y_obs), 
    alpha = 0.15, color = col[3],
    arrow=arrow(angle=15, length=unit(3, "mm"))) +
  labs(title = "Overestimating surprisal", 
       subtitle="schematic illustration", y="RT",x="surprisal") +
  ylim(NA,620) +
  # xkcd::xkcdaxis(range(df_modified$x),c(300,620)) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank(), axis.line = element_line(),
        text=element_text(size=11))
ggsave(
  paste0(figsdir, "sketch.pdf"),
  width = 5, height = 4.5)
```


# Sampling analysis


```{r options}
#| code-fold: show
#| code-summary: Options

# Whether to sample weights or use weights equidistributed on the inverse CDF
get_weights <- "sample" # or "equidistributed_inv_CDF"
```

```{r plotting_functions}
# Plotting functions
prepare_to_plot <- function(df) {
  df %>%
    mutate(across(contains("variance"), ~log(sqrt(.))
                  #, .names = "logsqrt{col}" # uncomment to not modify in place
                  )) %>%
    tidyr::pivot_longer(
      cols = matches("empirical|theoretical")) %>%
    tidyr::separate(name, into = c("moment", "type"), sep = "_") %>%
    mutate(across(where(is.character), forcats::as_factor)) %>%
    mutate(surprisal = -log(normalized_weight)) %>%
    arrange(desc(type))
}

make_plot <- function(df) {
  df_prepared <- prepare_to_plot(df)
  # rename "empirical" -> "simulated"
  df_prepared$type <- df_prepared$type %>%
    recode_factor(empirical = "simulated")
  df_prepared %>%
    ggplot(aes(
      x = surprisal, y = value,
      fill = type, color = type,
      shape = type, size = type, alpha = type)) +
    geom_point(alpha = 0.5) +
    facet_grid(
      rows = vars(moment), scale = "free", switch = "y",
      labeller = as_labeller(
        c(mean = "mean #samples", variance = "log(sd #samples)"))) +
    scale_color_manual(
      values = c(simulated = "darkgray", theoretical = "darkblue")) +
    scale_shape_manual(
      values = c(simulated = 4, theoretical = 16)) +
    scale_size_manual(
      values = c(simulated = 3, theoretical = 1)) +
    scale_alpha_manual(values = c(simulated = .5, theoretical = .8)) +
    theme(#legend.position = c(.2,.8),
          legend.background = element_rect(
            linewidth = 0, fill = alpha("white", 0.5), color = alpha("white", 0.5)),
          # legend.direction="horizontal",
          strip.placement = "outside",
          legend.title = element_blank(),
          axis.title.y = element_blank())
}
```

```{r generate_plots}
#| code-fold: true

suffix <- if (get_weights == "sample") "" else "_equidistant"

swor <- readr::read_csv(
  paste0(datadir, "sampling_simulated/plot_swor_runtime", suffix, "_df.csv"),
  show_col_types = FALSE)
swr <- readr::read_csv(
  paste0(datadir, "sampling_simulated/plot_swr_runtime", suffix, "_df.csv"),
  show_col_types = FALSE)
p_swor <- make_plot(swor)
p_swr <- make_plot(swr)

p_swor +
  labs(title = "Runtime of guessing without replacement",
       subtitle = "Pareto-distributed weights")
ggsave(
  filename = paste0(figsdir, "pareto", suffix, "_swor.pdf"),
  height = 4, width = 5)

p_swr +
  labs(title = "Runtime of guessing with replacement",
       subtitle = "Pareto-distributed weights")
ggsave(
  filename = paste0(figsdir, "pareto", suffix, "_swr.pdf"),
  height = 4, width = 5)

wrap_plots(
    p_swr + labs(subtitle = "Guessing with replacement"),
    p_swor + labs(subtitle = "Guessing without replacement"),
    guides = "collect") & #theme(legend.position = "right") &
  plot_annotation(
    title = "Runtime mean and variance of guessing algorithms",
    subtitle = "Pareto-distributed weights")
ggsave(
  filename = paste0(figsdir, "pareto", suffix, "_swr_swor", ".pdf"),
  height = 4, width = 7)

```

